---
title: Use of a calibration model for count data to predict cancer incidence in areas without cancer registry
subtitle: Prediction of cancer incidence using cancer registry and hospitalization data
author:
date: "8 February 2018"
output:
  html_document:
    fig_width: 10
    fig_height: 6
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
    theme: cosmo
---

In this document, we illustrate the main features of the `CalibInc` R
package through the example of estimating LOP cancer incidence in men
for the 96 French metropolitan districts over the 2007-2011
period. The predictions are made using registry incidence data from 14
districts and hospitalization data for all the districts.
This exemple served as an illustration of the method in the article
*"For a sound use of healthcare data in epidemiology: evaluation of a
calibration model for count data with application to prediction of
cancer incidence in areas without cancer registry"*.

> **WARNING**
>
> *In order to comply with diffusion constrains,*
> *the numbers of cancer and hospitalizations in the datasets*
> *were slightly altered, while keeping the same value and district*
> *variability of the H/C ratio, as in the real data.*


Installing CalibInc
====================

Use [devtools](https://github.com/hadley/devtools) to install the
latest version of `CalibInc` from Github (may take time):

```{r,echo=T,eval=F}
require(devtools)
install_github("echatignoux/CalibInc")
```

After installation, the package can be loaded into R.
```{r,echo=T,eval=T}
library(CalibInc)
```

Using CalibInc
==============
The use of `CalibInc` package is illustrated through the example of
the prediction of LOP cancer incidence in men.
The code can also be applied to the (altered) datasets for NHL
cancer incidence in women (datasets `mhlw.CalibSet` and
`mhlw.alldist`) which has significant variability in the H/C ratio.

*Note*: In addition to `CalibInc`, the following code needs the
`Hmisc`, `lme4` and `gridExtra` packages to be installed.


## Cancer incidence and hospitalization data sets ##

The dataset `lopm.CalibSet` comprise data on LOP cancer in
men over the 2007-2011 period for the 14 districts covered by a
cancer registry. The number of cancer incident cases (variable `C`)
and corresponding numbers of hospitalizations (variable `H`) for LOP
cancer in men are tabulated by 5 years age groups (from 0-5 to 90+,
variable `age` corresponds to the central age of each age group) and
district (variable `dist`). In order to diminish the number of age
groups with no cancer, the number of incident cases in lower ages
groups were aggregated.

```{r,echo=T}
library(tidyverse)
data(lopm.CalibSet)
lopm.CalibSet
```

The dataset `lopm.alldist` comprise hospitalization data for LOP cancer in
men in all French district over the 2007-2011 period. The data are
tabulated by 5 years age groups (from 0-5 to 90+, variable `age`
corresponds to the central age of the age group) and district (variable
`dist`). The variable `py` corresponds to the number of person years
and `WSP` stands for the World Standard Population of 1960 (Segi standard).

```{r,echo=T}
data(lopm.alldist)
lopm.alldist
```


## Evaluation of the calibration model ##
### Estimation of model parameters ###

In the calibration model (2.1), the mean H/C ratio (exp(f)) is modelled with a
restricted cubic spline with one knot at the median age (median over incident
cases) and boundary knots placed at 5 years above and below the minimum and maximum age
respectively (over all cases).

```{r,echo=T}
library(splines)
library(Hmisc)## Used to calculate weighted quantiles
k<-wtd.quantile(lopm.CalibSet$age,weights=lopm.CalibSet$C,p=0.5)%>%as.numeric
form.calib<-substitute(H~offset(log(C))+
                         ns(age,knots=k, Boundary.knots = range(age)+c(5,-5))+
                         (1|dist),
                       list(k=k))
form.calib<-as.formula(form.calib)
```

The model can be evaluated with the `glmer`function from the `lme4`
package:
```{r,echo=T}
library(lme4)
mod.calib<-glmer(form.calib,data=lopm.CalibSet%>%filter(C>0),family="poisson", nAGQ =20)
summary(mod.calib)
```
We can have a quick look at the mean H/C ratio:
```{r,echo=T}
age<-unique(lopm.CalibSet$age)
dtMeanHC<-data_frame(C=1,H=1,age=age)
## Predict H/C mean relation (i.e. predictions response with no random effects)
dtMeanHC$pred<-predict(mod.calib,newdata=dtMeanHC,re.form=NA,type="response")
## Display in a graph
qplot(data=dtMeanHC,age,pred,geom="line")+xlab("Age")+ylab("Mean H/C ratio")

```
The ratio has a U shape and is always greater that one (more hospitalisations than incident
cases).


### Model diagnostics ###

Traditional diagnostic plots may be used to check the consistency
between the model and the data. Note that residuals and fitted values
includes eventual random effects.

```{r,echo=T}
## Build table with deviance residuals conditional on random effects and fitted values
diag.mod<-lopm.CalibSet%>%filter(C>0)
diag.mod<-mutate(diag.mod,
                 res=residuals(mod.calib,type="deviance"),
                 fit=predict(mod.calib,type="link"),
                 y=qqnorm(res,plot=F)$x)
## Bissectrice line
bis<-geom_abline(intercept=0,slope=1)

## Display traditionnal diagnotics plots in a 2x2 layout
library(gridExtra)
grid.arrange(
  qplot(data=diag.mod,y,res)+bis+xlab("Theoretical Quantiles")+ylab("Sample Quantiles")+bis+
    ggtitle("Normal Q-Q plot"),
  qplot(data=diag.mod,fit,res)+xlab("Fitted values")+ylab("Residuals")+
    ggtitle("Resids vs. linear pred."),
  qplot(data=diag.mod,res,fill=I("white"),colour=I("black"),bins=15)+xlab("Residuals")+
    ggtitle("Histogram of residuals"),
  qplot(data=diag.mod,exp(fit),H)+xlab("Fitted values")+ylab("Response")+bis+
    ggtitle("Response vs. Fitted Values"),
  ncol=2)
```
The residuals behave reasonably, although the QQ plot may
suggest slight under-dispersion. This can be accounted for by
estimating the GLMM model with a PQL routine (via the `glmmPQL` function
from the `MASS` package):

```{r,echo=T}
mod.calibPQL<-MASS::glmmPQL(nobars(form.calib),random=~1|dist,
                            data=lopm.CalibSet%>%filter(C>0),family="poisson")
mod.calibPQL$sigma
```
The PQL estimation confirms a small under-dispersion in the data.


# Prediction of incidence #

Once the calibration model fit, predictions are done with the
`CalibInc` function. This function takes the calibration model as an
argument (`mod`) and requires a prediction table `pred` containing
the proxy of incidence (hospitalisation data in this application) and
the other variables used in the calibration model (only `age` and
`dist` in this application). `CalibInc` has a  `weight` argument used to weight
row predictions, and an `aggregate` argument used to aggregate the
predictions according to a combination of variables given as a formula
(see below).

The output is a `data_frame` with predicted incidence `pred` and
associated standard error `se`.

## Number of incident cases, crude and standardized rates ##


### 1. By age and district ###

Total number of incident cancer cases by age and district over the
2007-2011 period is predicted with:

```{r,echo=T}
pred.dist.age.lopm<-CalibInc(mod.calib,pred=lopm.alldist,aggregate=~age+dist)
pred.dist.age.lopm

```

Specific age-incidence rates by districts are obtained using the
`weight` argument, with weights equal to the inverse of the
number of person year by age in the district (times 100,000):

```{r,echo=T}
cr.dist.age.lopm<-CalibInc(mod.calib,pred=lopm.alldist,aggregate=~age+dist,
                           w=100000/lopm.alldist$py)
cr.dist.age.lopm

```

### 2. By district ###

Total number of incident cancer cases  by district over the
2007-2011 period is predicted with:

```{r,echo=T}
pred.dist.lopm<-CalibInc(mod.calib,pred=lopm.alldist,aggregate=~dist)
pred.dist.lopm

```

Crude incidence rates by districts are obtained using the
`weight` argument, with weights corresponding to the inverse of the
number of person year in the district (times 100,000):

```{r,echo=T}
lopm.alldist<-group_by(lopm.alldist,dist)
lopm.alldist<-mutate(lopm.alldist,w.cr=100000/sum(py))
cr.dist.lopm<-CalibInc(mod.calib,pred=lopm.alldist,aggregate=~dist,w=w.cr)
cr.dist.lopm
```

Similarly, age standardized rate by district can be obtained by using
a weight corresponding to the ratio between standard population figures
and the number of person year by age and district:

```{r,echo=T}
lopm.alldist$w.asr<-lopm.alldist$WSP/lopm.alldist$py
asr.dist.lopm<-CalibInc(mod.calib,pred=lopm.alldist,aggregate=~dist,w=w.asr)
asr.dist.lopm
```

### 3. For the whole territory ###

The total number of incident for cases over 2007-2011 for France as a
whole is predicted with:

```{r,echo=T}
pred.fr.lopm<-CalibInc(mod.calib,pred=lopm.alldist,aggregate=~1)
pred.fr.lopm

```

and crude incidence rates obtained using the
`weight` argument, with individual weights corresponding to the inverse of the
total number of person year (times 100,000):

```{r,echo=T}
lopm.alldist$w.cr.fr<-100000*1/sum(lopm.alldist$py)
cr.fr.lopm<-CalibInc(mod.calib,pred=lopm.alldist,aggregate=~1,w=w.cr.fr)
cr.fr.lopm
```

Finally, age standardized rate for France are obtained by using
a weight corresponding to the ratio between standardized population
and the number of person year by age:

```{r,echo=T}
lopm.alldist<-group_by(lopm.alldist,age)
lopm.alldist<-mutate(lopm.alldist,w.asr.fr=WSP/sum(py))
asr.fr.lopm<-CalibInc(mod.calib,pred=lopm.alldist,aggregate=~1,w=w.asr.fr)
asr.fr.lopm

```


## Prediction of SIR ##

The following code provides example of how to use `CalibInc` to
calculate SIR in each French districts.

### 1. Calculate expected incidence by district ###

The expected number of cases in a district is calculated as the sum of
age-incidence rate for France times the total district's person years.

```{r,echo=T}
## Calculate french age-incidence rate with weight equal to
## the inverse of total number of person-years for the age group
lopm.alldist<-group_by(lopm.alldist,age)
lopm.alldist<-mutate(lopm.alldist,w.cr.fr.age=1/sum(py))
cr.fr.age.lopm<-CalibInc(mod.calib,pred=lopm.alldist,aggregate=~age,w=w.cr.fr.age)
## Add the rate to individual table
cr.fr.age.lopm<-select(cr.fr.age.lopm,age,pred)
cr.fr.age.lopm<-rename(cr.fr.age.lopm,ar=pred)
lopm.alldist<-left_join(lopm.alldist,cr.fr.age.lopm,by="age")
## Sum by district
SIR<-group_by(lopm.alldist,dist)
SIR<-summarise(SIR,E=sum(py*ar))
SIR
```

### 2. Add predicted number of incident cases by districts and standard errors ###

```{r,echo=T}
SIR<-left_join(SIR,pred.dist.lopm)
SIR
```

### 3. Calculate the SIR and add prediction intervals thanks to `LogNormPI` function ###

```{r,echo=T}
SIR<-mutate(SIR,SIR=pred/E,se.SIR=se/E)
SIR<-LogNormPI(SIR,pred=SIR,se=se.SIR)
SIR
```

### 4. Represent the predicted SIR on a map ###

The `ggMap` function can be use to represent the predicted SIR on a map:
```{r,echo=T}
SIR<-mutate(SIR,id=dist,NS=!(low>1|up<1))
ggMap(data=SIR,
      SIR,color= "RdYlGn",rev=T,ns=NS,limits=c(0.7,1.3))+
  ggtitle("SIR for LOP cancer incidence in men, 2007-2011")


```

## Comparison with PQL estimates ##

As we saw, slight under-dispersion is present in the data. We can
account for the under-dispersion when calculating the variances of the
prediction by using the `glmmPQL` fit in the `mod` argument of
`CalibInc`.

```{r,echo=T}
## Calculate french age-incidence rate
cr.fr.age.lopm.PQL<-CalibInc(mod.calibPQL,pred=lopm.alldist,aggregate=~age,w=w.cr.fr.age)
cr.fr.age.lopm.PQL<-select(cr.fr.age.lopm.PQL,age,pred)
cr.fr.age.lopm.PQL<-rename(cr.fr.age.lopm.PQL,ar.PQL=pred)
## Add the rate to individual table
lopm.alldist<-left_join(lopm.alldist,cr.fr.age.lopm.PQL,by="age")
## Sum by district to get expected counts
SIR.PQL<-group_by(lopm.alldist,dist)
SIR.PQL<-summarise(SIR.PQL,E=sum(py*ar.PQL))
## Add predicted number of incident cases by districts and standard errors
pred.dist.lopm.PQL<-CalibInc(mod.calibPQL,pred=lopm.alldist,aggregate=~dist)
SIR.PQL<-left_join(SIR.PQL,pred.dist.lopm.PQL)
## Calculte SIR and prediction intervals
SIR.PQL<-mutate(SIR.PQL,SIR=pred/E,se.SIR=se/E)
SIR.PQL<-LogNormPI(SIR.PQL,pred=SIR,se=se.SIR)
SIR.PQL
```

The resulting predictions and standard errors are very close to the
one obtained with the `glmer` function, as can
be seen in the caterpillar graph below. In this graph, the districts
are ordered by increasing SIR values and bars represent 95% PI.

```{r,echo=T}
## Regroup glmer and glmmPQL predicted incidence
SIR.compare<-bind_rows(
  mutate(SIR,Method="glmer"),
  mutate(SIR.PQL,Method="glmmPQL"))
## Arrange districts by increasing SIR
SIR<-arrange(SIR,SIR)
SIR$rank<-1:nrow(SIR)
SIR.compare<-left_join(SIR.compare,select(SIR,dist,rank))
SIR.compare<-mutate(SIR.compare,rank=rank-0.4*(Method=="glmer"))
## Plot
qplot(data=SIR.compare,rank,SIR,colour=Method)+geom_pointrange(aes(ymin=low,ymax=up))+
  xlab("Districts")+theme(axis.text.x=element_blank(),axis.ticks.x=element_blank())
```

Only small differences were expected here as dispersion was low. However,
in case of marked dispersion, the use of `glmmPQL` is recommended.


















